Different forms of uncertainty differently affect the evolution of social learning

### 2022-07-17 ms.tex
further Model sec revisions

Evolution, then, acts as an optimization routine to identify which heritable trait
values are most rewarded under in different environmental uncertainty contexts,
defined by different uncertainty parameter settings
(Table~\ref{tab:uncertaintyParameters}).  The main outcome variable we analyze in
this model is therefore the value of agents' heritable trait, namely whether or not
an agent, $i$, learns socially, denoted $s_i$. $s_i$ is a binary variable: it is 1
if the agent is a social learner, and 0 otherwise. We will aggregate $s_i$ over all
agents and simulation trials to understand the affects of different uncertainty
parameter settings. We will support conclusions we make from observing patterns in
the outcomes with additional measures of the average time it takes for the
population to either all have $s_i = 1$ or all have $s_i = 0$, and the expected
payoffs in different contexts among populations that are initialized to all have
$s_i = 1$ or to all have $s_i = 0$. These outcome measures will be used to predict
the evolutionarily-favored strategy (to learn socially or not), the number of
generations it will take for a population to evolve the optimal strategy, and the
expected payoffs in populations performing the optimal strategy, in several specific
uncertainty contexts. To understand how different varieties of uncertainty interact
to affect the evolution of social learning, we systematically created an array of
simulated environments by systematically varying the uncertainty parameters.

@@

All agents perform individual, trial-and-error learning over their lifespan.
This learning is guided based on undirected, softmax search which drives
agents to explore all behavioral options in the selection set when 
the agent is more uncertain which is optimal, and to exploit the most profitable
behaviors when the agent is more certain which are most beneficial. Selection
for or against social learners occurs when agents are chosen to reproduce 
with replacement, with probability
of reproduction biased by net payoffs in each generation.

Several agent-level variables are required to operationalize individual and social
learning

@@

\paragraph{Initialization}

At generation start, one of the behaviors is set to 
yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. At each time step, agents perform behavior $b$ 
with softmax-weighted probability
\begin{equation}
  \Pr(\text{Agent $i$ performs behavior $b$}) = 
    \frac{e^{\beta \ledger}}{\sum_{b=1}^B e^{\beta \ledger}},
\end{equation}
\noindent
where $\beta$ is a parameter that adjusts how frequently agents perform 
behaviors with high expected payoffs (larger $\beta$) versus how frequently
agents explore alternative behaviors (smaller $\beta$)~\cite{McElreath2005}. 
Agents probabilistically recieve a payoff of 0 or 1 depending on whether the
bandit paid off. Behavior payoffs are represented as a draw from a 
Bernoulli distribution with mean $\pi_b$, so $\pi_b$ is also the mean payoff from
behavior $b$.  For concreteness we denote this probabilistic payoff
$\bandit$. On performing behavior $b$, agent $i$ updates the
corresponding \emph{behavior count} by 1, $c_{ib} \leftarrow c_{ib} + 1$, and then
the expected payoffs calculated for that behavior are updated with
exponentially-weighted averaging
\begin{equation}
  \ledger \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}}.
\end{equation}
\noindent
This procedure continues for $L$ time steps within each generation
(Figure~\ref{fig:schematic}B).

\subsubsection{Intragenerational dynamics}

To guide behavior selection, agent $i$ counts how many times it
performed each behavior $b$, denoted $c_{ib}$. Agents use this count to 
update the mean payoff they have obtained from behavior $b$, denoted $\ledger$.
$\ledger$ is initialized to 0 for all $b$ at model start for
all agents (Figure~\ref{fig:schematic}A), and at 
generation start for asocial learners. Social
learning agents' $\ledger$ is initialized as that of its teacher, who is selected
through performance-biased oblique learning; $c_ib$ is set to 1 for all social
learners $i$ and behaviors $b$ (Figure~\ref{fig:schematic}C). Agent $i$'s
accumulated payoffs from performing several behaviors over their lifetime of $L$
time steps is denoted $\pi_{i}$. If an agent $i$ is a social learner, then we write
$s_i = 1$, otherwise the agent is an asocial learner and $s_i = 0$.

\subsubsection{Intergenerational dynamics}

In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off (Figure~\ref{fig:schematic}C). 
$N$ reproducers are selected with replacement over $N$ independent draws, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
A child inherits its parent's social learning trait $s_i$ without mutation.
A social learner child with $s_i = 1$ learns from a teacher from its parent's
generation, including possibly their parent (oblique learning). 
A child selects its teacher by first selecting $N_T$ prospective
teachers from the population completely at random, then selecting the one with
highest net payoffs $\pi_i$ as their teacher, with ties broken randomly ($N_T =
5$ in the main text; \mt{Other values will be tested in the supplement}). While it
is somewhat arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally guaranteed.
Social learners adopt their chosen teacher's ledger of expected payoffs $\ledger$,
and set all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$.  Asocial
learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and $\ledger = 0$ 
(Figure~\ref{fig:schematic}C). The
newly initialized population then begins behavior selection and payoff accumulation
for another $L$ time steps. The evolutionary process continues until $\sum_i s_i
= 1$ or $\sum_i s_i = 0$ \mt{I believe all trials will eventually end like this, so
in the next set of model runs I will not allow early termination before one of these
conditions are met}.

>>>>

Evolution, then, acts as an optimization routine to identify which heritable trait
values are most rewarded under in different environmental uncertainty contexts,
defined by different uncertainty parameter settings
(Table~\ref{tab:uncertaintyParameters}).  The main outcome variable we analyze in
this model is therefore the value of agents' heritable trait, namely whether or not
an agent, $i$, learns socially, denoted $s_i$. $s_i$ is a binary variable: it is 1
if the agent is a social learner, and 0 otherwise. We will aggregate $s_i$ over all
agents and simulation trials to understand the affects of different uncertainty
parameter settings. We will support conclusions we make from observing patterns in
the outcomes with additional measures of the average time it takes for the
population to either all have $s_i = 1$ or all have $s_i = 0$, and the expected
payoffs in different contexts among populations that are initialized to all have
$s_i = 1$ or to all have $s_i = 0$. These outcome measures will be used to predict
the evolutionarily-favored strategy (to learn socially or not), the number of
generations it will take for a population to evolve the optimal strategy, and the
expected payoffs in populations performing the optimal strategy, across several specific
uncertainty contexts. To understand how different varieties of uncertainty interact
to affect the evolution of social learning, we systematically created an array of
simulated environments by systematically varying the uncertainty parameters.

@@

All agents perform individual, trial-and-error learning over their lifespan.
This learning is guided based on the softmax search algorithm which drives
agents to explore all behavioral options in the selection set when 
the agent is more uncertain which is optimal, and to exploit the most profitable
behaviors more frequently when the agent is more certain which are most beneficial. 
To do softmax searching, agents track payoffs acquired from each behavior.
(introduce agent attributes)
When agents are initialized at the start of the model, all agents are initialized
with all $\pi_{ib} = 0$. At the start of generations other than the first, 
$\pi_i$ will be initialized with all $\pi_{ib} = 0$ for asocial learners ($s_i = 0$)
or with a teacher's $\pi_{ib}$. 
Agents are thus endowed with dynamic attributes to track payoffs and calculate the
softmax weights, and with an instruction of whether child agents learn socially
or not. Details of how these attributes are updated over time is explained below
with the rest of the model dynamics.

@@

\paragraph{Initialization}
At generation start, one of the behaviors is set to 
yield expected payoff $\pihigh$, with the other $B-1$ behaviors yielding
expected payoffs $\pilow$. At each time step, agents perform behavior $b$ 
with softmax-weighted probability
\begin{equation}
  \Pr(\text{Agent $i$ performs behavior $b$}) = 
    \frac{e^{\beta \ledger}}{\sum_{b=1}^B e^{\beta \ledger}},
\end{equation}
\noindent
where $\beta$ is a parameter that adjusts how frequently agents perform 
behaviors with high expected payoffs (larger $\beta$) versus how frequently
agents explore alternative behaviors (smaller $\beta$)~\cite{McElreath2005}. 
Agents probabilistically recieve a payoff of 0 or 1 depending on whether the
bandit paid off. Behavior payoffs are represented as a draw from a 
Bernoulli distribution with mean $\pi_b$, so $\pi_b$ is also the mean payoff from
behavior $b$.  For concreteness we denote this probabilistic payoff
$\bandit$. On performing behavior $b$, agent $i$ updates the
corresponding \emph{behavior count} by 1, $c_{ib} \leftarrow c_{ib} + 1$, and then
the expected payoffs calculated for that behavior are updated with
exponentially-weighted averaging
\begin{equation}
  \ledger \leftarrow \ledger + \frac{\bandit - \ledger}{c_{ib}}.
\end{equation}
\noindent
This procedure continues for $L$ time steps within each generation
(Figure~\ref{fig:schematic}B).

\paragraph{Intragenerational dynamics}
To guide behavior selection, agent $i$ counts how many times it
performed each behavior $b$, denoted $c_{ib}$. Agents use this count to 
update the mean payoff they have obtained from behavior $b$, denoted $\ledger$.
$\ledger$ is initialized to 0 for all $b$ at model start for
all agents (Figure~\ref{fig:schematic}A), and at 
generation start for asocial learners. Social
learning agents' $\ledger$ is initialized as that of its teacher, who is selected
through performance-biased oblique learning; $c_ib$ is set to 1 for all social
learners $i$ and behaviors $b$ (Figure~\ref{fig:schematic}C). Agent $i$'s
accumulated payoffs from performing several behaviors over their lifetime of $L$
time steps is denoted $\pi_{i}$. If an agent $i$ is a social learner, then we write
$s_i = 1$, otherwise the agent is an asocial learner and $s_i = 0$.

\paragraph{Intergenerational dynamics}
In between generations agents first reproduce via asexual haploid reproduction; 
then new child agents learn from the
previous generation if they are social learners; and finally all agents from the
previous generation then die off (Figure~\ref{fig:schematic}C). 
$N$ reproducers are selected with replacement over $N$ independent draws, 
biased by performance:
\begin{equation}
  \Pr(\text{Agent $i$ is chosen to reproduce in one draw}) = \frac{\pi_i}{\sum_{i=1}^N \pi_i}.
\end{equation}
\noindent
A child inherits its parent's social learning trait $s_i$ without mutation.
A social learner child with $s_i = 1$ learns from a teacher from its parent's
generation, including possibly their parent (oblique learning). 
A child selects its teacher by first selecting $N_T$ prospective
teachers from the population completely at random, then selecting the one with
highest net payoffs $\pi_i$ as their teacher, with ties broken randomly ($N_T =
5$ in the main text; \mt{Other values will be tested in the supplement}). While it
is somewhat arbitrary to first subset $N_T$, this is a conservative choice that
represnts the fact that access to the entire population is not generally guaranteed.
Social learners adopt their chosen teacher's ledger of expected payoffs $\ledger$,
and set all $c_{ib} = 1$ so expected payoffs remain within $[0, 1]$.  Asocial
learners ($s_i = 0$) are initialized with all $c_{ib} = 0$ and $\ledger = 0$ 
(Figure~\ref{fig:schematic}C). The
newly initialized population then begins behavior selection and payoff accumulation
for another $L$ time steps.

\paragraph{Stopping conditions} The evolutionary process continues until $\sum_i s_i
= 1$ or $\sum_i s_i = 0$.

\mt{I believe all trials will eventually end like this, so
in the next set of model runs I will not allow early termination before one of these
conditions are met}.
